# Data Visualisation {#PM6}

```{r tidyr6, include = FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=50), tidy = FALSE)
```

You have successfully downloaded and summarised your NatureCounts dataset. In this chapter we will demonstrate how to do some basic visulizations with plots and maps. 

##Plotting {#PM6.1}

Plotting your data is a powerful way to detect patterns and make results stand out. Recall in [Chapter 2](#Package2.2) you installed the [tidyverse](https://www.tidyverse.org/) package, which included `ggplot2` for data visulizations. You are encouraged to learn more about this function by reviewing [Cookbook for R](http://www.cookbook-r.com/Graphs/). We also recommend you print a copy of the RStudio [Data Visualization](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) cheat sheet as a reference document.

At the end of [Chapter 5](#Manip5) you were exploring fall migration monitoring data collected at [Vaseux Lake Bird Observatory](https://www.birdscanada.org/birdmon/default/datasets.jsp?code=CMMN-DET-VLBO), British Columbia. Let's continue using this dataset for our plotting example and download the data for 2015-present. 

First, lets apply our previously aquired skills to download the filtered dataset and zero-fill for GRCA, while also keep some extra variables. 

```{r VLBO, eval = FALSE}

VLBO<-nc_data_dl(collections="CMMN-DET-VLBO", years=c(2015, NA), username = "sample", info = "test plot")

GRCA<-format_zero_fill(VLBO, species= 15900, by=c("SamplingEventIdentifier", "survey_year", "survey_month", "survey_day"))

```

First, we are interested if there are any noticible patterns in migration timing. For this, we will use the [add date and day-of-year helper function](https://birdstudiescanada.github.io/naturecounts/reference/format_dates.html) to add two new columbs to the dataframe. 

```{r doy, eval = FALSE}

GRCA_dates<-format_dates(GRCA)

```

Now we can plot raw counts (y-axis) for each day-of-year (x-axis). 

```{r plot1VLBO}

ggplot(data = GRCA_dates) + 
  geom_point(aes(x = doy, y = ObservationCount))

```

What you will notice is that there isn't an obvious peak in migration for this species, but a realtively constant number of individuals counted throughout the season. 

Next, we are interested in visually examining the mean number of migrant GRCA each year, to see if there are any noticible changes over time. First, we need to summarise the data: 

```{r VLBOsum, eval = FALSE}

#use this shortcut function to calculate the standard error
se <- function(x) sd(x) / sqrt(length(x))

GRCA_year<-GRCA %>% group_by(survey_year) %>% summarise (MeanObs=mean(ObservationCount), SEObs = se(ObservationCount)) %>% 
  mutate (yrmin=MeanObs+SEObs, yrmax=MeanObs-SEObs)

```

Now we can create the plot: 

```{r plot2VLBO}

ggplot(data = GRCA_year) + 
  geom_pointrange(aes(x = survey_year, y=MeanObs, ymin=yrmin, ymax=yrmax))
  
```

##Mapping {#PM6.2}

There is a comprehensive tutorial online for [Mapping Observation](https://birdstudiescanada.github.io/naturecounts/articles/articles/mapping-observations.html). The materials are not repeated here. We encourage you to walk through the online examples prior to   

